---
Author: Jean Rene Ndeki
Date: Monday, August 11, 2014
Output: html_document
Title: Peer Assessment Project 1
output: html_document
---
#Peer Assessment Project 1#
######By Jean Rene Ndeki######
######Monday, August 11, 2014######

```{r}
setwd('C:/Users/Jean-Rene/Desktop/Coursera/7. Reproducible Research/Project 1/Data')
```

####I. Loading and preprocessing the data####

**I.1. Load the data (i.e. read.csv())**

```{r}
activity <- read.csv("activity.csv")
```

**I.2. Process/transform the data (if necessary) into a format suitable for your analysis**

```{r}
tsp_Ag <-  aggregate(activity[, 'steps'], by=list(activity$date), sum, na.rm = TRUE)
colnames(tsp_Ag) <- c('Date', 'Steps')
```

####II. What is mean total number of steps taken per day?####

For this part of the assignment, you can ignore the missing values in the dataset.

**II.1. Make a histogram of the total number of steps taken each day**

```{r, message=FALSE,echo=TRUE}
library(ggplot2)
qplot(tsp_Ag$Steps,geom="histogram",fill=I("light green"), main="Plot 1. Daily Total Number Steps",xlab="Total steps",ylab="Frequency")
```

**II.2. Calculate and report the mean and median total number of steps taken per day**

```{r}
mean(tsp_Ag$Steps)
median(tsp_Ag$Steps)
```

####III. What is the average daily activity pattern?####

**III.1. Make a time series plot (i.e. type = "l") of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all days (y-axis)**

```{r}
time <- formatC(activity$interval/100, 2, format = "f")
activity$datetime <- as.POSIXct(paste(activity$date, time), format = "%Y-%m-%d %H.%M",tz = "GMT")
activity$time <- format(activity$datetime, format = "%H:%M:%S")
activity$time <- as.POSIXct(activity$time, format = "%H:%M:%S")
```

```{r}
tspt_Ag <-  aggregate(activity[, 'steps'], by=list(activity$time), sum, na.rm = TRUE)
colnames(tspt_Ag) <- c('Interval', 'Steps')
```

```{r, echo=TRUE}
library(scales)
ggplot(tspt_Ag, aes(Interval, Steps)) + 
    geom_line() + 
    ggtitle('Plot 2. Total Number of Steps Taken per Time Interval') +
    xlab("Time Interval") + 
    ylab("Average Number of Steps") + 
    scale_x_datetime(labels = date_format(format = "%H:%M"))
```

**III.2. Which 5-minute interval, on average across all the days in the dataset, contains the maximum number of steps?**

```{r}
max5 <- which.max(tspt_Ag$Steps)
format(tspt_Ag[max5, "Interval"], format = "%H:%M")
```

####IV. Imputing missing values####
Notice there is a number of days/intervals where there are missing values (coded as NA). The presence of missing days may introduce bias into some calculations or summaries of the data

**IV.1. Calculate and report the total number of missing values in the dataset (i.e. the total number of rows with NAs)**

```{r}
summary(activity$steps)
```


**IV.2. Devise a strategy for filling in all of the missing values in the dataset. The strategy does not need to be sophisticated. For example, you could use the mean/median for that day, or the mean for that 5-minute interval, etc.**

```{r, message=FALSE}
library(Hmisc)
act_imp <- activity
act_imp$steps <- with(act_imp, impute(steps, mean))
```

**IV.3. Create a new dataset that is equal to the original dataset but with the missing data filled in.**

```{r}
tsp_imp_Ag <-  aggregate(act_imp[, 'steps'], by=list(act_imp$date), sum, na.rm = TRUE)
colnames(tsp_imp_Ag) <- c('Date', 'Steps')
```

**IV. Histogram, Mean, Median, Comparison, and Impact**

**IV.4.1. Make a histogram of the total number of steps taken each day.**

```{r, message=FALSE, echo=TRUE}
qplot(tsp_imp_Ag$Steps, geom = "histogram", fill=I("blue"), main="Plot 3. Total Number Daily Steps - Imputed",xlab = "Total steps", ylab = "Frequency")
```

**IV.4.2. Calculate and report the mean and median total number of steps taken per day**

```{r}
tsp_imp <- tapply(act_imp$steps, act_imp$date, sum)
```

```{r}
mean(tsp_imp)
median(tsp_imp)
```

**IV.4.3. Do these values differ from the estimates from the first part of the assignment?** 

*Yes. The values differ from the assignment first part estimates. However, notice the imputed data set mean and median are equal.*

**IV.4.4. What is the impact of imputing missing data on the estimates of the total daily number of steps?**

```{r}
mtsp<-mean(tsp_Ag$Steps)
metsp<-median(tsp_Ag$Steps)
mtspimp<-mean(tsp_imp)
metspimp<-median(tsp_imp)
```

```{r}
mtspimp-mtsp
metspimp-metsp
```

*The impact of imputing missing data on the estimates of the total daily number of steps is the imputed missing data set mean and median are higher than those of the dataset with missing values*

####V. Are there differences in activity patterns between weekdays and weekends?####

For this part the weekdays() function may be of some help here. Use the dataset with the filled-in missing values for this part.

**V.1. Create a new factor variable in the dataset with two levels - "weekday" and "weekend" indicating whether a given date is a weekday or weekend day.**

```{r}
wkday <- function(date) {
    if (weekdays(date) %in% c("Saturday", "Sunday")) {
        return("Weekend")
    } else {
        return("Weekday")
    }
}
wkdays <- sapply(act_imp$datetime, wkday)
act_imp$wkday <- as.factor(wkdays)

mSteps <- tapply(act_imp$steps, interaction(act_imp$time, act_imp$wkday), mean, na.rm = TRUE)

wkdaydf <- data.frame(time = as.POSIXct(names(mSteps)), mSteps = mSteps, 
                               wkday = as.factor(c(rep("Weekday", 288), rep("Weekend", 288))))
```

**V.2. Make a panel plot containing a time series plot (i.e. type = "l") of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all weekday days or weekend days (y-axis). The plot should look something like the following, which was creating using simulated data.**

```{r, echo=TRUE}
library(scales)
ggplot(wkdaydf, aes(time, mSteps)) + 
    geom_line() + 
    xlab("Time of day") + 
    ylab("Mean number of steps") + 
    ggtitle('Plot 4. Panel Weekday and Weekend Step Average Number - Imputed') +
    scale_x_datetime(labels = date_format(format = "%H:%M")) + 
    facet_grid(. ~ wkday)
```
